Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-07 Manuscript/manuscript.Rmd
results paragraph 1 edits

## DATA CONSTRUCTION FOR UNBIASED COMPARISONS BETWEEN RESOLUTION-SPECIFIC RANDOM FOREST CLASSIFIERS
<!--## CELL LINE AND DATA [??? What data?] RESOLUTION SPECIFIC MODEL [??? What model?] CONSTRUCTION [??? Why?] !!!created new header-->

In total, we analyzed [??? Very boring] publically available Hi-C data for both the GM12878 and K562 cell lines, at four resolutions including 10 kb, 25 kb, 50 kb, and 100 kb. [??? But why? Problem is not defined] Given the varying levels of data resolution, we created a resolution-specific approach toward data constuction used for downstream modelling (Figure 1). This allowed us an unbiased way to compare performances of random forest classifiers between varying resolutions of Hi-C data. [??? Problem is not defined]

![**Model Construction for TAD boundary classification.** A subsection of the contact matrix for chromosome 14 on GM12878 [@durand2016juicer] at 100kb resolution shows the appearance of TAD boundaries. To account for the uncertainty in TAD boundary location, each boundary point was flanked on both sides by resolution-specific genomic regions (bins). The response vector $Y$ was constructed by defining $Y=1$ if a bin flanks a TAD boundary, and $Y=0$ otherwise. <!--The spatial relationship between positional coordinates of various functional genomic annotations, obtained from the UCSC Genome Browser Database, together with the genomic bins were then used to define the feature space of the models represented by the data matrix $\textbf{X} = \{X_1, X_2, ..., X_p\}$, where $p$ was the complete set of functional annotations.-->](Figures/figure1_model_construction_V3.png)

@@

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models [?? In methods you say you do "feature selection" and then test model types. This implies the opposite ??]. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites), across resolution. We found that for random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the cell-line specific transcription factor binding sites, and was consistent across resolutions (Figure 5; Supplementary Figure 4).

@@

We then sought to determine which of the transcription factor binding sites were most influential in predicting TAD boundaries. We found that the cell line specific binding sites associated with SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were given consistently high predictive importance, across all resolutions [?? Is this after elastic net? Elastic net distributes coefficients between correlated variables and it may affect variable importance ??] (Figure 7; Supplementary Figure 6). The SMC3, CTCF, and RAD21 binding sites are known factors associated with the cohesin protein complex, which is hypothesized to be what creates loops in the DNA  [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, ZNF143 and YY1 are known structural regulators of enhancer-promoter interactions that also contribute to looping of the linear genome [@weintraub2017yy1; @gonzalez2017znf143]. With the identification of these factors known to be enriched at TAD boundaries, we are able to confirm the validity of our model building process. Additionally, we were also able to identify some lesser known cell line specific transcription factors as predictive of TAD boundaries including the MYC-associated zing-finger protein (MAZ), jun-D proto-oncogene (JUND), and ELF1. These transcription factor binding sites tended to be given higher predictive importance at lower resolutions. In fact, at lower resolutions, the predictive importance was found to be much more evenly dispersed across the top transcription factor binding sites, than for higher resolutions. This is again suggestive of the fact that the spatial relationship between functional annotations, specifically transcription factor binding sites, and TAD boundaries changes at different Hi-C data resolutions.

>>>>

## DEVELOPING A RESOLUTION-SPECIFIC APPROACH FOR DATA CONSTRUCTION
<!--## CELL LINE AND DATA [??? What data?] RESOLUTION SPECIFIC MODEL [??? What model?] CONSTRUCTION [??? Why?] !!!created new header-->

We developed a resolution-specific approach toward constructing data necessary for TAD boundary prediction. We illustrate the proposed method of construction using the following example (Figure 1). Given in-situ Hi-C data sequenced at 100 kb resolution, we binned the linear genome into 100 kb intervals. Using the Arrowhead TAD calling algorithm [@durand2016juicer], unique TAD boundaries were identified and flanked on either side by the corresponding bin width, $b$, for a total flanking width of $2b$ (i.e. if $b$=100 kb, the total flanked width of each TAD would be 200 kb). We defined the response vector **Y** as a variable that indicated whether a genomic bin overlapped with a flanked TAD boundary ($Y=1$) or not ($Y=0$). In total, we applied our method of data construction for four resolutions including 10 kb, 25 kb, 50 kb, and 100 kb.

![**An illustration of the proposed approach for resolution-specific data construction used for TAD boundary prediction.** <!--A subsection of the contact matrix for chromosome 14 on GM12878 [@durand2016juicer] at 100kb resolution shows the appearance of TAD boundaries. To account for the uncertainty in TAD boundary location, each boundary point was flanked on both sides by resolution-specific genomic regions (bins). The response vector $Y$ was constructed by defining $Y=1$ if a bin flanks a TAD boundary, and $Y=0$ otherwise.-->](Figures/figure1_model_construction_V3.png)

@@

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites), across resolution. We found that for random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the cell-line specific transcription factor binding sites, and was consistent across resolutions (Figure 5; Supplementary Figure 4).

@@

We then sought to determine which of the transcription factor binding sites were most influential in predicting TAD boundaries. We found that the cell line specific binding sites associated with SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were given consistently high predictive importance, across all resolutions (Figure 7; Supplementary Figure 6). The SMC3, CTCF, and RAD21 binding sites are known factors associated with the cohesin protein complex, which is hypothesized to be what creates loops in the DNA  [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, ZNF143 and YY1 are known structural regulators of enhancer-promoter interactions that also contribute to looping of the linear genome [@weintraub2017yy1; @gonzalez2017znf143]. With the identification of these factors known to be enriched at TAD boundaries, we are able to confirm the validity of our model building process. Additionally, we were also able to identify some lesser known cell line specific transcription factors as predictive of TAD boundaries including the MYC-associated zing-finger protein (MAZ), jun-D proto-oncogene (JUND), and ELF1. These transcription factor binding sites tended to be given higher predictive importance at lower resolutions. In fact, at lower resolutions, the predictive importance was found to be much more evenly dispersed across the top transcription factor binding sites, than for higher resolutions. This is again suggestive of the fact that the spatial relationship between functional annotations, specifically transcription factor binding sites, and TAD boundaries changes at different Hi-C data resolutions.
